An Experimental Analysis of the Influence of the Precision of Parameters in an Artificial Neural Network

### 2025-11-12 paper/main.tex
weiter

\section{Experimental Design}

The effect of numerical precision on training and inference is evaluated
through five experiment groups run on standard image classification tasks. The
main dataset MNIST\footnote{LeCun, Cortes, and Burges:
	\url{http://yann.lecun.com/exdb/mnist/}.} ($28\times 28$ grayscale images of
handwritten digits, train/test split $60.000$/$10.000$, 10 classes, examples
shown in Figure~\ref{fig:mnistSample}) is used to determine precision effects
on accuracy and runtime under a simple workload, while an after-training
rounding experiment is performed on FashionMNIST\footnote{Xiao, Rasul, and
	Vollgraf: \url{https://github.com/zalandoresearch/fashion-mnist}.} ($28\times
28$ grayscale images of clothing items, train/test split $60.000$/$10.000$, 10
classes) so the impact of using coarse value sets is better observable.\\

@@

@@

\subsection*{Accuracy Across Bitwidths}
Figure~\ref{fig:modelAccTrainAndInf} shows how the accuracy changed after the
models were trained on the different precision settings. Keep in mind that the
y-axis is not starting from zero, so the absolute height differences are
amplified visually, though the delta is only around $.1\%$. Re-training the
models and re-evaluating them has also lead to the accuracies being
``reversed'', meaning the FP64 model had the least accuracy. It is safe to
assume that the model accuracy has neither decreased, nor increased.
\begin{figure}[H]
	\centering
	\includegraphics[width=0.85\textwidth]{figures/trainingAndInference.png}
	\caption{Model accuracy evaluated after training with different precisions.}\label{fig:modelAccTrainAndInf}
\end{figure}

@@

\subsection*{Automatic Mixed Precision and Gradient Clipping}
Automatic Mixed Precision (AMP) accelerates deep learning training by using
both FP16 and FP32 data types reducing memory usage and increasing
computational speed. However, FP16's limited dynamic range can cause numerical
instability, especially during backpropagation. To address this, Gradient
Clipping is used to cap the magnitude of gradients (either globally or per
element) preventing extreme updates that could hinder convergence. Together,
AMP and gradient clipping enable efficient, stable training by balancing
performance and numerical precision.

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.85\textwidth]{figures/118_accuracy.png}
	\caption{Test Accuracy vs.\ number of decimal digits retained in weights.}\label{fig:118_acc}
\end{figure}

Figure~\ref{fig:118_acc} shows how rounding the decimal part affects the test
accuracy. For just one decimal the model has a maximum of 19 different weights
at its disposal, which makes the model have only 10\% accuracy, random guessing
with a 10 class problem. Two decimals perform significantly better. At 70\%
accuracy the model gets close to the full precision 88\% accuracy with 32 bit
floating point. Making and educated guess on distribution such as
Figure~\ref{fig:118_2} seems to be a good heuristic for this example. At just 3
decimals, the full 32 bit precision is reached. Thus we can conclude that with
just 10 bit for representing 3 decimals, no bits for the integer part and one
additional bit for the sign, only 11 bit must be stored to have the best
performing model for inference.

@@

% This work confirms that for relatively simple tasks like MNIST classification,
% significant reductions in parameter precision can be achieved without
% compromising performance. Our future efforts will extend this analysis to more
% complex datasets (e.g., CIFAR-10, ImageNet) and deeper architectures, and
% explore advanced quantization-aware training techniques.

>>>>

\section{Experimental Design}
The effect of numerical precision on training and inference is evaluated
through five experiment groups run on standard image classification tasks. The
main dataset MNIST\footnote{LeCun, Cortes, and Burges:
	\url{http://yann.lecun.com/exdb/mnist/}.} ($28\times 28$ grayscale images of
handwritten digits, train/test split $60.000$/$10.000$, 10 classes, examples
shown in Figure~\ref{fig:mnistSample}) is used to determine precision effects
on accuracy and runtime under a simple workload, while an after-training
rounding experiment is performed on FashionMNIST\footnote{Xiao, Rasul, and
	Vollgraf: \url{https://github.com/zalandoresearch/fashion-mnist}.} ($28\times
28$ grayscale images of clothing items, train/test split $60.000$/$10.000$, 10
classes) so the impact of using coarse value sets is better observable.\\

@@

For the MNIST experiments (i) end-to-end precision, (ii) timing, (iii)
inference-only precision reduction, and (iv) automatic mixed precision (AMP)
with gradient clipping, we use 5-fold cross-validation on the training set with
random equal splits. Unless noted otherwise, figures report fold means, and no
test set tuning is performed. In the end-to-end runs, every floating point
tensor appearing inside the training loop (parameters, activations, gradients,
and optimizer state) is kept at the target precision during training and
evaluation. The timing experiment reuses these runs to measure mean per-epoch
training time and mean evaluation time. For inference-only reduction, we first
train a baseline model in FP64 and then do inference on copies cast to FP32 and
FP16 without any further optimization (so no \textit{retraining} on the lower
precision).\\

The AMP experiment is run on MNIST with gradient clipping, comparing an AMP
setup (FP16 computations with FP32 master weights) against a pure FP32
baseline. We try different global-norm and per-element value-clipping
thresholds to check stability and report test accuracy and training/inference
times. We also run a one-off rounding experiment on FashionMNIST (batch size
$256$, FP32 training): after training, only the stored weights are rounded to
$k$ decimal places ($k{=}1\ldots14$) and inference remains FP32. Inputs and
intermediate computations stay at full precision. We report test accuracy as a
function of $k$ and note the impact on storage.

@@

\subsection*{Accuracy Across Bitwidths}
Figure~\ref{fig:modelAccTrainAndInf} shows how the accuracy changed after the
models were trained on the different precision settings. Keep in mind that the
y-axis is not starting from zero, so the absolute height differences are
amplified visually, though the delta is only around $.1\%$. Retraining the
models and re-evaluating them has also lead to the accuracies being
``reversed'', meaning the FP64 model had the least accuracy. It is safe to
assume that the model accuracy has neither decreased, nor increased.
\begin{figure}[H]
	\centering
	\includegraphics[width=0.85\textwidth]{figures/trainingAndInference.png}
	\caption{Model accuracy evaluated after training with different precisions.}\label{fig:modelAccTrainAndInf}
\end{figure}

@@

\subsection*{Automatic Mixed Precision and Gradient Clipping}
Automatic Mixed Precision (AMP) improves deep learning training by using both
FP16 and FP32 data types reducing memory usage and increasing computational
speed. However, FP16's limited dynamic range can cause numerical instability,
especially during backpropagation. To address this, Gradient Clipping is used
to cap the magnitude of gradients (either globally or per element) preventing
extreme updates that could hinder convergence. Together, AMP and gradient
clipping enable efficient, stable training by balancing performance and
numerical precision.

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.85\textwidth]{figures/118_accuracy.png}
	\caption{Test accuracy vs.\ number of decimal digits retained in weights.}\label{fig:118_acc}
\end{figure}

Figure~\ref{fig:118_acc} shows how rounding the decimal part affects the test
accuracy. For just one decimal the model has a maximum of 19 different weights
at its disposal, which makes the model have only 10\% accuracy, random guessing
with a 10 class problem. Two decimals perform significantly better. At 70\%
accuracy the model gets close to the full precision 88\% accuracy with 32 bit
floating point. Making an educated guess on distribution such as
Figure~\ref{fig:118_2} seems to be a good heuristic for this example. At just 3
decimals, the full 32 bit precision is reached. Thus we can conclude that with
just 10 bit for representing 3 decimals, no bits for the integer part and one
additional bit for the sign, only 11 bit must be stored to have the best
performing model for inference.

@@
